---
title: "Commuter membership — fixed origin, destination, and departure hour"
date: "`r Sys.Date()`"
output:
  github_document: default
---

Reports the commuter membership of paper Sec. 7.8 (`sec:commuter`): 100 counterfactual
commuters, each holding a 30-ride monthly membership on one origin–destination pair at
one departure hour, priced under both travel-time models and both strikes.

Each commuter's trips on the pair are split into a **history** set, which defines the
route distribution the membership is priced against, and a **held-out** trip, randomized
into the 30 rides the member actually takes. Held-out trips are excluded from the
travel-time model fit, and both sides of the ledger pass through `similar_route()`, so the
sampler's length bias cannot show up as provider profit.

Run `Rscript run_commuter_membership.R` first; it saves
`../plot/commuter_membership_results.rds`, which this report reads.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 300)
RESULTS_RDS <- "../plot/commuter_membership_results.rds"
```

```{r load}
library(data.table)
library(ggplot2)
source("commuter_heatmap.R")

if (!file.exists(RESULTS_RDS)) {
  stop("Missing ", RESULTS_RDS, ". Run run_commuter_membership.R first.")
}
results <- readRDS(RESULTS_RDS)
str(results$settings)
```

## Membership performance

Returns are per member: the collected premium less realized overruns, as a percentage of
the fares the member generates over the month.

```{r summary-table}
knitr::kable(format_summary_table(results), align = "lcrrrrrrr")
```

```{r premium-share, fig.width = 7.5, fig.height = 3.4}
plot_premium_share(results)
```

```{r profit-distribution, fig.width = 7.5, fig.height = 4}
plot_profit_distribution(results)
```

## Breakage

A member pays the premium up front but need not take all 30 rides. Columns give the share
of the month's rides actually taken.

```{r usage-lines, fig.width = 7.5, fig.height = 4.5}
usage_grid <- prepare_usage_grid(results)
plot_usage_lines(usage_grid, k_factor = 1)
```

```{r usage-heatmap-profit, fig.width = 8.5, fig.height = 6}
plot_usage_heatmap(usage_grid, "mean_profit")
```

```{r usage-heatmap-return, fig.width = 8.5, fig.height = 6}
plot_usage_heatmap(usage_grid, "mean_pct_return")
```

## Data availability

The number of origin–destination pairs able to supply a commuter shrinks quickly with the
history size, which bounds how far the information axis can be pushed on this data.

```{r eligible-pairs}
knitr::kable(unique(results$summary[, .(
  `History size` = history_size,
  `Eligible OD pairs` = n_eligible_pairs
)]))
```

```{r export-figures}
dir.create("../plot", showWarnings = FALSE, recursive = TRUE)
ggsave("../plot/commuter_premium_share.png", plot_premium_share(results),
       width = 7.5, height = 3.4, dpi = 300)
ggsave("../plot/commuter_profit_distribution.png", plot_profit_distribution(results),
       width = 7.5, height = 4, dpi = 300)
ggsave("../plot/commuter_usage_profit.png", plot_usage_heatmap(usage_grid, "mean_profit"),
       width = 8.5, height = 6, dpi = 300)
ggsave("../plot/commuter_breakage_lines.png", plot_usage_lines(usage_grid, k_factor = 1),
       width = 7.5, height = 4.5, dpi = 300)
file.copy(
  "../plot/commuter_breakage_lines.png",
  "../../latex/preprint/img/commuter-breakage.png",
  overwrite = TRUE
)
```
