---
title: "Flat-membership abuse stress test — loss ratio & break-even loading"
author: "Mohamad Elmasri"
date: "`r Sys.Date()`"
output:
  github_document: default
---

Stress-tests an **unrestricted flat membership**: a single cap `Kbar` covers any
route at any time, so the rider's marginal price of distance is **zero** and a
rational holder substitutes toward the longest covered trips. Abuse is therefore
selected on **trip length** (the dimension the flat price fails to condition on).

Two design fixes over the earlier grid:

- **Metric.** We report an insurance **loss ratio** `LR = claims / premium`, where
  `claims = sum_i (P_i - Kbar)^+` over the rider's `N` trips and `premium = N * Rbar`.
  Because the premium is a constant, the cell-mean loss ratio **is** the break-even
  premium **loading** `theta` (the multiple of the fair premium the provider must
  charge to survive that abuse level). `theta > 1` means the product is underwater.
- **Global quantile.** The severity cut is taken from the **whole pool**, not each
  rider's 40-trip sample, which removes the unstable high-`q` cells.

`Rbar, Kbar` are calibrated empirically from the same simulator, so under **no abuse**
the loss ratio is `~1` by construction (a sanity check, below). The rational-abuser
corner `(M/N, q) -> (1, 0.9)` is reported explicitly as the **security level**.

> NOTE: not executed here. Verify the `traveltimeCLT` helpers (`price`,
> `dependent_uniform`, `get_timeBin_x_edges`, `time_bins_readable`) resolve in your
> environment, and check the no-abuse row prints `theta ~ 1` before trusting the grid.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

```{r params}
library(traveltimeCLT)
library(data.table)

SEED              <- 1234L
RIDERS_PER_REPEAT <- 100L                       # counterfactual riders per cell
TRIPS_PER_RIDER   <- 40L                         # N rides per membership
K_FACTOR          <- 0.9                          # flat cap = 0.9 * mean expected price (10% discount)
                                                  #   set 1.0 for an undiscounted flat cap
Q_LEVELS <- seq(0.1, 0.9, by = 0.1)              # severity: draw abused trips from the upper-q of LENGTH
M_VALUES <- seq(0, TRIPS_PER_RIDER, by = TRIPS_PER_RIDER * 0.1)  # M/N = 0, 0.1, ..., 1.0 (incl. corner)

# calibration budget for the flat premium/cap (means, so a subsample is ample)
CAL_N    <- 1500L
CAL_REPS <- 100L

stopifnot(all(M_VALUES == floor(M_VALUES)), max(M_VALUES) <= TRIPS_PER_RIDER)
```

```{r helpers-engine}
## --- simulation engine, verbatim from the project's sensitive-test file --------
fill_timebin_edges <- function(unique_trips, timebin_x_edges) {
  dt <- copy(unique_trips)
  global_dt <- timebin_x_edges[timeBin == "Global"]
  na_idx <- which(is.na(dt$mean))
  cols <- c("timeBin", "mean", "sd", "frequency", "length", "ID")
  if (length(na_idx) > 0) {
    repl <- global_dt[dt[na_idx, .(linkID)], on = "linkID"]
    dt[na_idx, (cols) := repl[, mget(cols)]]
  }
  na.omit(dt)
}

simulate_trip_prices <- function(sampled, trips, timebin_x_edges) {
  unique_trips <- sampled[, .(n_rep = .N), .(tripID, timeBin)]
  unique_trips <- merge(unique_trips, trips[, .(tripID, linkID)], by = "tripID", allow.cartesian = TRUE)
  unique_trips <- merge(unique_trips, timebin_x_edges, by = c("timeBin", "linkID"), all.x = TRUE)
  unique_trips <- fill_timebin_edges(unique_trips, timebin_x_edges)

  travel <- unique_trips[, {
    reps <- n_rep[1]
    times <- numeric(reps)
    dist <- sum(length)
    for (i in seq_len(reps)) {
      times[i] <- sum(exp(mean + stats::qnorm(dependent_uniform(.N)) * sd))
    }
    .(time = times, distance = rep(dist, reps), index = seq_len(reps))
  }, by = .(timeBin, tripID)]

  out <- copy(sampled)
  out[, index := seq_len(.N), by = .(tripID, timeBin)]
  out <- merge(out, travel, by = c("tripID", "timeBin", "index"), all.x = TRUE)
  out[, real_price := price(time, distance)[, 1]]
  out
}
```

```{r helpers-flat}
## --- per-trip pool: length (selection axis) + the trip's own start time bin ----
build_pool_stat <- function(trips) {
  trips[, .(
    distance = sum(distance_meters),   # meters; the length dimension abuse selects on
    timeBin  = timeBin[1]              # trip's true starting bin (not randomized)
  ), by = tripID]
}

## --- calibrate the flat membership price: Kbar (single cap) and Rbar (fair premium)
## Kbar = K_FACTOR * mean_pool E[P];   Rbar = E_pool[(P - Kbar)^+]
## Estimated from the SAME simulator so the no-abuse loss ratio is ~1 by construction.
calibrate_flat_membership <- function(stat, trips, timebin_x_edges,
                                      k_factor = K_FACTOR,
                                      n_cal = CAL_N, reps = CAL_REPS) {
  cal_ids <- sample(stat$tripID, min(n_cal, nrow(stat)))
  cal     <- stat[tripID %in% cal_ids]
  sampled <- cal[, .(trip_idx = seq_len(reps)), by = .(tripID, timeBin)]  # reps draws per trip
  priced  <- simulate_trip_prices(sampled, trips[, .(tripID, linkID)], timebin_x_edges)

  EP   <- priced[, .(EP = mean(real_price)), by = tripID]      # per-trip expected price
  Kbar <- k_factor * mean(EP$EP)                                # single flat cap
  Rbar <- priced[, mean(pmax(real_price - Kbar, 0))]            # fair per-ride membership premium
  list(Kbar = Kbar, Rbar = Rbar, mean_EP = mean(EP$EP))
}

## --- the abuse grid: distance selection, GLOBAL quantile, loss-ratio metric -----
run_flat_abuse_grid <- function(stat, trips, timebin_x_edges, Kbar, Rbar,
                                M_values = M_VALUES, q_levels = Q_LEVELS,
                                riders = RIDERS_PER_REPEAT, N = TRIPS_PER_RIDER) {
  premium <- N * Rbar                                          # constant across riders/cells

  build_cell <- function(q) {
    cut        <- as.numeric(stats::quantile(stat$distance, probs = q, names = FALSE))  # GLOBAL
    abuse_pool <- stat[distance >= cut]
    if (nrow(abuse_pool) == 0L) stop("Empty abuse pool at q = ", q)
    CJ(rider = seq_len(riders), M = M_values)[, {
      abuse_ids  <- if (M > 0L)
        abuse_pool$tripID[sample.int(nrow(abuse_pool), M, replace = TRUE)] else integer(0)
      normal_ids <- stat$tripID[sample.int(nrow(stat), N - M, replace = TRUE)]
      .(trip_idx = seq_len(N), tripID = c(abuse_ids, normal_ids), q = q)
    }, by = .(rider, M)]
  }

  sampled <- rbindlist(lapply(q_levels, build_cell))
  sampled <- merge(sampled, stat[, .(tripID, timeBin)], by = "tripID", all.x = TRUE)

  priced <- simulate_trip_prices(sampled, trips[, .(tripID, linkID)], timebin_x_edges)

  rider <- priced[, {
    claims <- sum(pmax(real_price - Kbar, 0))                  # sum_i (P_i - Kbar)^+
    expend <- sum(real_price)
    .(loss_ratio = claims / premium,                           # LR; cell-mean = break-even loading theta
      profit     = premium - claims,
      pct_return = (premium - claims) / expend * N)            # legacy metric, for comparability
  }, by = .(q, M, rider)]

  rider[, abuse_rate := M / N]
  rider[, .(
    theta        = mean(loss_ratio),          # break-even premium loading (= mean loss ratio)
    se_theta     = stats::sd(loss_ratio) / sqrt(.N),
    p_underwater = mean(loss_ratio > 1),      # share of riders whose claims exceed premium
    mean_return  = mean(pct_return),
    se_return    = stats::sd(pct_return) / sqrt(.N)
  ), by = .(abuse_rate, q)][order(abuse_rate, q)]
}

to_wide <- function(results_long, value.var) {
  dcast(results_long, abuse_rate ~ q, value.var = value.var)
}
```

```{r data-setup}
set.seed(SEED)
trips <- fread("../data/trips.csv")
names(trips)[c(2, 3, 5, 7, 8)] <- c(
  "tripID", "entry_time", "duration_secs", "distance_meters", "linkID"
)
trips$speed   <- exp(trips$logspeed)
trips$timeBin <- time_bins_readable(trips$entry_time)

timebin_x_edges <- get_timeBin_x_edges(
  tripID   = trips$tripID,
  linkId   = trips$linkID,
  length   = trips$distance_meters,
  timeBin  = trips$timeBin,
  duration = trips$duration_secs
)
setnames(timebin_x_edges, c("timeBin", "linkId"), c("timeBin", "linkID"))

stat <- build_pool_stat(trips)
cat("Pool trips:", nrow(stat), " | median length (m):", stats::median(stat$distance), "\n")
```

```{r calibrate}
set.seed(SEED)
cal <- calibrate_flat_membership(stat, trips, timebin_x_edges)
cat(sprintf("Flat cap Kbar = %.2f  (=%.2f x mean E[P] = %.2f)\n",
            cal$Kbar, K_FACTOR, cal$mean_EP))
cat(sprintf("Fair membership premium Rbar = %.3f per ride;  N*Rbar = %.2f collected\n",
            cal$Rbar, TRIPS_PER_RIDER * cal$Rbar))
```

```{r run-grid}
set.seed(SEED)
results_long <- run_flat_abuse_grid(stat, trips, timebin_x_edges,
                                    Kbar = cal$Kbar, Rbar = cal$Rbar)
results_long
```

```{r sanity-no-abuse}
# Under no abuse (M/N = 0), q is irrelevant and theta should scatter around ~1.
baseline <- results_long[abuse_rate == 0]
cat("No-abuse row: theta range =", round(range(baseline$theta), 3),
    "| mean =", round(mean(baseline$theta), 3), "(should be ~1)\n")
print(baseline[, .(q, theta, p_underwater)])
```

```{r security-level}
# Rational-abuser corner: zero marginal length-price -> longest covered trips.
corner <- results_long[abuse_rate == max(abuse_rate) & q == max(q)]
cat(sprintf(
  "Security level (M/N=%.1f, q=%.1f): break-even loading theta = %.2f (%.0f%% of riders underwater)\n",
  corner$abuse_rate, corner$q, corner$theta, 100 * corner$p_underwater))
cat("=> to survive this abuse level the provider must charge", round(corner$theta, 2),
    "x the fair premium; if theta is large the flat product is effectively uninsurable.\n")
```

```{r wide-tables}
theta_wide        <- to_wide(results_long, "theta")            # break-even loading surface
underwater_wide   <- to_wide(results_long, "p_underwater")     # P(claims > premium) surface
theta_wide
underwater_wide
```

```{r export, eval=FALSE}
dir.create("plot_abuse", showWarnings = FALSE, recursive = TRUE)
fwrite(results_long,     "plot_abuse/flat_abuse_long.csv")
fwrite(theta_wide,       "plot_abuse/flat_abuse_theta.csv")
fwrite(underwater_wide,  "plot_abuse/flat_abuse_underwater.csv")
```

```{r heatmap, fig.width=5.5, fig.height=5, dpi=300}
library(ggplot2)
library(scales)

# grayscale, journal-ready: darker = worse (higher loading); solvent cells (theta<=1) outlined.
plot_theta <- function(res, cap = 3, label_size = 2.4) {
  d <- as.data.frame(res)
  d$.txt <- ifelse(pmin(d$theta, cap) > (1 + 0.55 * (cap - 1)), "grey95", "grey10")
  ggplot(d, aes(q, abuse_rate, fill = theta)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_tile(data = subset(d, theta <= 1), fill = NA, colour = "black", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.2f", theta), colour = .txt), size = label_size) +
    scale_colour_identity() +
    scale_fill_gradient(low = "grey97", high = "grey20", name = expression(theta),
                        limits = c(min(d$theta), cap), oob = scales::squish,
                        guide = guide_colourbar(barwidth = unit(3, "mm"),
                                                barheight = unit(35, "mm"))) +
    coord_fixed(ratio = 1, expand = FALSE) +
    labs(title = "Break-even premium loading",
         x = expression(paste("Severity quantile ", italic(q), " (trip length)")),
         y = expression(paste("Abuse fraction ", italic(M), "/N"))) +
    theme_minimal(base_size = 9) +
    theme(panel.grid = element_blank(),
          plot.title = element_text(hjust = 0.5),
          axis.text  = element_text(colour = "black"))
}
plot_theta(results_long)
```

```{r note-model-contrast, eval=FALSE}
## To recover the population-vs-trip-specific *model-risk* contrast on top of this,
## compute Kbar/Rbar from each model's PRICED premium instead of the empirical
## calibrator, e.g. fit <- traveltimeCLT(trips, model); pt <- predict(fit, trips_record);
## then set Kbar/Rbar from request_K/request_R against the flat cap. Claims stay
## simulator-based (the "truth"); only the priced premium changes with the model.
## For flat/length abuse this is second-order (distance is priced exactly by both
## models via C2*delta), so the headline result above is model-robust.
```
