---
title: "Membership abuse stress test — 100 riders × 40 trips"
author: "Mingze Li"
date: "`r Sys.Date()`"
output:
  github_document: default
---

Stress-tests **discount membership** (10% cap, `K = 0.9P`) by simulating **100 counterfactual
riders** with **40 trips** each. A fraction **M₁/40** of trips are drawn from the upper-**q**
quantile of the trip-length distribution on a per-rider test sample; **q** controls severity,
**M₁/40** controls frequency. When **M₁ = 0**, no trip is drawn from the tail and **q** is
irrelevant—the top row is nine independent Monte-Carlo replicates of the no-abuse baseline.

Based on `table4(2).rmd` / paper Sec. Membership Abuse (`tab:abuse-return`, `tab:abuse-sd`).
Runs the same grid under **trip-specific** and **population** travel-time models (`traveltimeCLT`).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r params}
library(traveltimeCLT)
library(data.table)

SEED <- 1234L
N_RIDERS <- 1L          # 100 counterfactual riders (10 repeats × 10 riders)
N_REPEATS <- 1L
RIDERS_PER_REPEAT <- 100L
TRIPS_PER_RIDER <- 40L    # M_1 <= 40 abusive trips per rider
K_FACTOR <- 1.0
Q_LEVELS <- seq(0.1, 0.9, by = 0.1)   # severity quantile (columns in paper tables)
M_VALUES <- seq(0, TRIPS_PER_RIDER * 0.9, by = TRIPS_PER_RIDER * 0.1)  # M_1/40 = 0, 0.1, …, 0.9
stopifnot(N_RIDERS == N_REPEATS * RIDERS_PER_REPEAT)
stopifnot(all(M_VALUES == floor(M_VALUES)))
Q_LEVELS <- c(0)
M_VALUES <- c(0)
```

```{r helpers}
fill_timebin_edges <- function(unique_trips, timebin_x_edges) {
  dt <- copy(unique_trips)
  global_dt <- timebin_x_edges[timeBin == "Global"]
  na_idx <- which(is.na(dt$mean))
  cols <- c("timeBin", "mean", "sd", "frequency", "length", "ID")
  if (length(na_idx) > 0) {
    repl <- global_dt[dt[na_idx, .(linkID)], on = "linkID"]
    dt[na_idx, (cols) := repl[, mget(cols)]]
  }
  na.omit(dt)
}

simulate_trip_prices <- function(sampled, trips, timebin_x_edges) {
  unique_trips <- sampled[, .(n_rep = .N), .(tripID, timeBin)]
  unique_trips <- merge(unique_trips, trips[, .(tripID, linkID)], by = "tripID", allow.cartesian = TRUE)
  unique_trips <- merge(unique_trips, timebin_x_edges, by = c("timeBin", "linkID"), all.x = TRUE)
  unique_trips <- fill_timebin_edges(unique_trips, timebin_x_edges)

  travel <- unique_trips[, {
    reps <- n_rep[1]
    times <- numeric(reps)
    dist <- sum(length)
    for (i in seq_len(reps)) {
      times[i] <- sum(exp(mean + stats::qnorm(dependent_uniform(.N)) * sd))
    }
    .(time = times, distance = rep(dist, reps), index = seq_len(reps))
  }, by = .(timeBin, tripID)]

  out <- copy(sampled)
  out[, index := seq_len(.N), by = .(tripID, timeBin)]
  out <- merge(out, travel, by = c("tripID", "timeBin", "index"), all.x = TRUE)
  out[, real_price := price(time, distance)[, 1]]
  out
}

run_membership_abuse_grid <- function(
  trips_stat,
  trips,
  timebin_x_edges,
  M_values = M_VALUES,
  q_levels = Q_LEVELS,
  n_repeats = N_REPEATS,
  riders_per_repeat = RIDERS_PER_REPEAT,
  trips_per_rider = TRIPS_PER_RIDER
) {
  pressure_data <- CJ(
    repeat_time = seq_len(n_repeats),
    q = q_levels,
    rider = seq_len(riders_per_repeat),
    M = M_values
  )
  pressure_data <- pressure_data[, .(trip_idx = seq_len(trips_per_rider)), by = .(repeat_time, q, rider, M)]

  cat(
    "Sampling", n_repeats, "repeats ×", riders_per_repeat, "riders ×",
    length(M_values), "abuse levels ×", length(q_levels), "q levels …\n"
  )

  sampled <- pressure_data[, {
    id <- sample.int(nrow(trips_stat), trips_per_rider, replace = FALSE)
    test <- trips_stat[id]
    train <- trips_stat[-id]
    assignments <- .SD[, {
      q_cut <- as.numeric(stats::quantile(test$distance, probs = q, names = FALSE))
      abuse_pool <- test[distance >= q_cut]
      if (M > 0L && nrow(abuse_pool) == 0L) {
        stop("Empty abuse pool for q = ", q, "; increase test sample or lower q.")
      }
      abuse_ids <- if (M > 0L) {
        abuse_pool$tripID[sample.int(nrow(abuse_pool), M, replace = TRUE)]
      } else {
        integer(0)
      }
      normal_ids <- test$tripID[sample.int(trips_per_rider, trips_per_rider - M, replace = TRUE)]
      tripID <- c(abuse_ids, normal_ids)
      Rt <- rep(mean(train$Rt), trips_per_rider)
      Kt <- rep(mean(train$Kt), trips_per_rider)
      timeBin <- sample(
        c("EveningNight", "EveningRush", "Weekday", "MorningRush", "Weekendday"),
        trips_per_rider,
        replace = TRUE
      )
      .(trip_idx = seq_len(trips_per_rider), tripID, Rt, Kt, timeBin)
    }, by = .(q, rider, M)]
    assignments
  }, by = .(repeat_time)]

  cat("Simulating travel times …\n")
  priced <- simulate_trip_prices(sampled, trips[, .(tripID, linkID)], timebin_x_edges)

  cat("Aggregating rider-level returns …\n")
  rider_returns <- priced[, {
    profit <- sum(Rt) + sum(pmin(Kt - real_price, 0))
    expand <- sum(real_price)
    pct_return <- profit / expand * trips_per_rider
    .(pct_return = pct_return, profit = profit, expand = expand)
  }, by = .(M, q, repeat_time, rider)]

  rider_returns[, abuse_rate := M / trips_per_rider]
  rider_returns[, .(
    mean_return = mean(pct_return, na.rm = TRUE),
    sd_return_se = stats::sd(pct_return, na.rm = TRUE) / sqrt(.N)
  ), by = .(abuse_rate, q)]
}

build_trips_stat <- function(trips, trips_record, model = c("trip-specific", "population")) {
  model <- match.arg(model)
  cat("Fitting", model, "model …\n")
  fit <- traveltimeCLT(trips, model)
  cat("Predicting travel times for user records …\n")
  pt <- predict(fit, trips_record)
  stat <- trips[, .(
    start_time = entry_time[1],
    distance = sum(distance_meters),
    duration_secs = max(entry_time) - min(entry_time),
    real_price = price(max(entry_time) - min(entry_time), sum(distance_meters))[, 1]
  ), tripID]
  stat[, Kt := request_K(pt, distance, discount_factor = K_FACTOR)]
  stat[, Rt := request_R(pt, start_time, start_time, distance, K = K_FACTOR, risk_free = 0, zeta = 0)]
  na.omit(stat)
}

to_wide_tables <- function(results_long) {
  return_dt <- dcast(results_long, abuse_rate ~ q, value.var = "mean_return", fun.aggregate = identity)
  setnames(return_dt, as.character(Q_LEVELS), paste0("q_", Q_LEVELS))
  sd_dt <- dcast(results_long, abuse_rate ~ q, value.var = "sd_return_se", fun.aggregate = identity)
  setnames(sd_dt, as.character(Q_LEVELS), paste0("q_", Q_LEVELS))
  list(return_dt = return_dt, sd_dt = sd_dt)
}

run_membership_abuse_scenario <- function(
  trips,
  trips_record,
  timebin_x_edges,
  model = c("trip-specific", "population")
) {
  model <- match.arg(model)
  cat("\n=== ", model, " model ===\n", sep = "")
  set.seed(SEED)
  trips_stat <- build_trips_stat(trips, trips_record, model)
  cat("Trips in pool:", nrow(trips_stat), "\n")
  results_long <- run_membership_abuse_grid(
    trips_stat = trips_stat,
    trips = trips,
    timebin_x_edges = timebin_x_edges
  )
  wide <- to_wide_tables(results_long)
  c(
    list(model = model, trips_stat = trips_stat, results_long = results_long),
    wide
  )
}
```

```{r data-setup}
set.seed(SEED)
trips <- fread("../data/trips.csv")
names(trips)[c(2, 3, 5, 7, 8)] <- c(
  "tripID", "entry_time", "duration_secs", "distance_meters", "linkID"
)
trips$speed <- exp(trips$logspeed)
trips$timeBin <- time_bins_readable(trips$entry_time)

trips_record <- trips[, .(
  entry_time = rep(entry_time[1], .N),
  timeBin,
  logspeed,
  tripID,
  linkID,
  distance_meters
), tripID]

timebin_x_edges <- get_timeBin_x_edges(
  tripID = trips$tripID,
  linkId = trips$linkID,
  length = trips$distance_meters,
  timeBin = trips$timeBin,
  duration = trips$duration_secs
)
setnames(timebin_x_edges, c("timeBin", "linkId"), c("timeBin", "linkID"))

cat("Loaded", uniqueN(trips$tripID), "trips for membership-abuse grid\n")
```

```{r trip-specific-simulation}
trip_specific <- run_membership_abuse_scenario(
  trips = trips,
  trips_record = trips_record,
  timebin_x_edges = timebin_x_edges,
  model = "trip-specific"
)
trip_specific$results_long
```

```{r trip-specific-tables}
return_dt <- trip_specific$return_dt
sd_dt <- trip_specific$sd_dt
return_dt
sd_dt
```

```{r population-simulation}
population <- run_membership_abuse_scenario(
  trips = trips,
  trips_record = trips_record,
  timebin_x_edges = timebin_x_edges,
  model = "population"
)
population$results_long
```

```{r population-tables}
return_dt_pop <- population$return_dt
sd_dt_pop <- population$sd_dt
return_dt_pop
sd_dt_pop
```

```{r no-abuse-row-check}
# When M_1/40 = 0, q should not matter (independent MC replicates of the same baseline).
for (scenario in list(trip_specific, population)) {
  baseline <- scenario$results_long[abuse_rate == 0, .(mean_return, sd_return_se, q)]
  cat("\n", scenario$model, " — no-abuse row: mean return range =",
      round(range(baseline$mean_return), 3),
      ", SE range =",
      round(range(baseline$sd_return_se), 3), "\n")
  print(baseline)
}
```

```{r export, eval=FALSE}
dir.create("plot_abuse", showWarnings = FALSE, recursive = TRUE)
fwrite(trip_specific$return_dt, "plot_abuse/return_dt_trip_specific.csv")
fwrite(trip_specific$sd_dt, "plot_abuse/sd_dt_trip_specific.csv")
fwrite(population$return_dt, "plot_abuse/return_dt_population.csv")
fwrite(population$sd_dt, "plot_abuse/sd_dt_population.csv")
saveRDS(
  list(
    trip_specific = trip_specific[c("return_dt", "sd_dt", "results_long")],
    population = population[c("return_dt", "sd_dt", "results_long")]
  ),
  "../plot/sensitive_test_results.rds"
)
```

```{r heatmaps, fig.width=5.5, fig.height=5, dpi=300}
library(ggplot2)
source("sensitive_test_heatmap.R")

ts_plots <- plot_scenario_heatmaps(trip_specific, "Trip-specific model")
ts_plots$return
```

```{r sd-heatmap, fig.width=5.5, fig.height=5.5, dpi=300}
ts_plots$sd
```

```{r combined-heatmap, fig.width=5.5, fig.height=5.5, dpi=300}
ts_plots$combined
```

```{r facet-heatmap, fig.width=11, fig.height=5.5, dpi=300}
ts_plots$facet
```

```{r population-heatmaps, fig.width=5.5, fig.height=5, dpi=300}
pop_plots <- plot_scenario_heatmaps(population, "Population model")
pop_plots$return
```

```{r population-sd-heatmap, fig.width=5.5, fig.height=5, dpi=300}
pop_plots$sd
```

```{r population-combined-heatmap, fig.width=5.5, fig.height=5.5, dpi=300}
pop_plots$combined
```

```{r population-facet-heatmap, fig.width=11, fig.height=5.5, dpi=300}
pop_plots$facet
```
