---
title: "Selective exercise in the commuter membership"
date: "`r Sys.Date()`"
output:
  github_document: default
---

Reports the selective-exercise section of the paper (`sec:selective-exercise`), which
follows the commuter membership of Sec. 7.8. The commuter holds 30 capped rides on a
fixed origin–destination pair at a fixed departure hour. Sec. 7.8 drops the rides that
are not taken **at random**, which makes unused rides pure profit. A rational holder
does not drop at random: they exercise the membership on the mornings the ride looks
expensive and use an outside option when it looks cheap.

The rider's private benefit on occasion *i* is the amount the cap saves them,
\((P_i - K)^+\), which is exactly the provider's payout, so rider objective and provider
loss are the same object and no separate utility model is needed. Foresight
\(\lambda \in [0,1]\) is the rank correlation between the rider's signal and the
realized fare: \(\lambda = 0\) reproduces Sec. 7.8, \(\lambda = 1\) is a rider who
always exercises on the worst-traffic days.

Nothing here is re-priced. Premiums, caps, and realized fares are read from the
commuter study's cached artefacts. Run `Rscript run_selective_exercise.R` first; it
saves `../plot/selective_exercise_results.rds`, which this report reads.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 300)
RESULTS_RDS <- "../plot/selective_exercise_results.rds"
```

```{r load}
library(data.table)
library(ggplot2)
source("selective_exercise_figures.R")

if (!file.exists(RESULTS_RDS)) {
  stop("Missing ", RESULTS_RDS, ". Run run_selective_exercise.R first.")
}
results <- readRDS(RESULTS_RDS)
curves <- results$curves
str(results$settings)
```

## Acceptance tests

These run inside `run_selective_exercise()` and hold up the whole run if any fails.
Test 2 is the structural one: taking all 30 rides pays the same regardless of the order
they were taken in. Test 7 is the regression test protecting the published Sec. 7.8
breakage result.

```{r tests}
knitr::kable(
  results$tests[, .(
    `#` = test, Check = description,
    Statistic = formatC(statistic, format = "g", digits = 4),
    Tolerance = formatC(tolerance, format = "g", digits = 4),
    Pass = pass
  )],
  align = "clrrc"
)
```

## Profit and loss ratio against utilization

The \(\lambda = 0\) line is the published breakage curve. Every line above it is what
the same month costs once the rider chooses which days to exercise. The claim to read
off the lower panel: the loss ratio is **linear** in utilization at \(\lambda = 0\) and
**concave** for \(\lambda > 0\), because the provider's exposure is front-loaded onto
the occasions the rider picks. The gap between the two is the cost of selective
exercise.

Open circles on the upper panel are the published Sec. 7.8 breakage curve. They sit on
the \(\lambda = 0\) line, which is acceptance test 7 by eye.

```{r curves, fig.width = 11, fig.height = 8}
plot_selective_panels(curves, results$reference_usage)
```

At full utilization every curve meets, whatever the foresight: taking all 30 rides
costs the provider the same total regardless of the order they are taken in. Selection
only moves *when* the exposure is paid, which is why it matters exactly to the extent
that the member does not use the whole month.

```{r surface, fig.width = 10, fig.height = 6.5}
plot_loss_ratio_surface(curves)
```

## Break-even utilization

\(u^*(\lambda)\) is the first utilization at which the loss ratio reaches one, linearly
interpolated between adjacent ride counts. `NA (>1)` means the product never breaks even
inside the month. Intervals are member-level nonparametric bootstrap percentiles,
resampling members rather than (member, \(\lambda\)) cells so the pairing induced by the
common random numbers survives.

Two loss ratios are carried through, and they answer different questions. The
**member-average** ratio averages each member's own payout-to-premium ratio, so every
member counts once and it describes the typical membership. The **pooled** ratio divides
total payout by total premium, so members count in proportion to what they pay; it is
the provider's book, and it crosses one exactly where mean profit crosses zero, which is
what Sec. 7.8 reports. The premium distribution is right skewed enough for the two to
part company — at full use under route conditioning at \(K = P\) the member average is
1.30 while the book is 1.08 and the median member is at 0.70 — so the paper headlines
the pooled version.

```{r break-even-table}
knitr::kable(format_break_even_table(results$break_even), align = "lccrrrr")
```

An interval is withheld where most resamples never break even inside the month, since
it would describe the minority that did rather than the quantity asked for. With 100
commuters the intervals are wide: the data caps the h = 8 cohort at 141 possible
commuters, so this axis cannot be sharpened by resampling more of it.

```{r break-even-figure, fig.width = 8, fig.height = 4}
plot_break_even(results$break_even)
```

## How good a forecaster the rider must be

Rather than estimating the rider's true foresight, which no data here identifies, the
question is inverted: at an assumed utilization \(u_0\), how good a forecaster must the
rider be before the product loses money?

```{r foresight-table}
knitr::kable(format_foresight_table(results$lambda_star), align = "lccrrr")
```

At \(u_0 = 1\) the answer is degenerate by construction — the whole month costs the same
under every \(\lambda\) — so that row reports whether the product is under water at full
use, not a foresight threshold.

## Export

```{r export-figures}
dir.create("../plot", showWarnings = FALSE, recursive = TRUE)
ggsave("../plot/selective_exercise_curves.png",
       plot_selective_panels(curves, results$reference_usage),
       width = 11, height = 8, dpi = 300)
ggsave("../plot/selective_exercise_surface.png", plot_loss_ratio_surface(curves),
       width = 10, height = 6.5, dpi = 300)
ggsave("../plot/selective_exercise_break_even.png", plot_break_even(results$break_even),
       width = 8, height = 4, dpi = 300)

# The paper carries the profit panel alone. Under the pooled normalization the loss
# ratio is the same curve rescaled -- Λ/Π = 1 − profit/Π with Π fixed in k and λ — so a
# second panel would repeat the first against a relabelled axis, and profit in dollars
# is directly comparable to the breakage figure of Sec. 7.8.
ggsave("../plot/selective_exercise_profit.png", plot_selective_profit(curves),
       width = 11, height = 4.5, dpi = 300)
file.copy(
  "../plot/selective_exercise_profit.png",
  "../../latex/preprint/img/selective-exercise.png",
  overwrite = TRUE
)
```
